Batson Challenges Using Historical Selection Data

### 2021-07-26 analysis/paper_draft.Rmd
added motivation for power prior; minor edits

This paper demonstrates a prototype software application that, with both contemporaneous and historical jury-selection data, attorneys can use in real time to estimate the probability that an attorney has used peremptory challenges to strike potential jurors because of juror race or gender, in violation of _Batson v. Kentucky_ (1986), and its progeny.

@@

In the US, once a group of individuals appear in court, this group is questioned by the parties' attorneys and the trial judge.  In turn, each individual may be (1) excused by the trial judge for cause; (2) selected for the jury; or (3) dismissed upon the exercise by one of the parties' attorneys of a peremptory challenge. The key feature of a peremptory challenge is that, with it, a party declares a prospective juror ineligible ("strikes" that juror) for a seat on the jury, but without the burden of explaining why. Each side is assigned a number of peremptory challenges and may use them to strike any potential juror without explanation. The number of peremptory challenges varies by jurisdiction, as does peremptory-challenge procedure, including the order of strikes and the number of peremptory strikes a party may exercise at a time [@nscs-jury2018; @Williams2017].

@@

Under _Batson v. Kentucky_ (1986)_, however, a party's exercise of a peremptory challenge violates the Equal Protection Clause of the US Constitution if motivated by a prospective juror's race, gender, or certain other protected characteristics. When raising a _Batson_ challenge during jury selection, the challenging party must typically establish a "prima facie case", that is, enough evidence to infer that a party struck a juror because of such juror's race, gender, or other protected status. Once established, the party exercising the strike bears the burden of producing a nondiscriminatory reason for that strike.  Thereafter, the challenger bears the burden of persuading the trial judge that the race or gender, not the proffered nondiscriminatory reason, was more likely than not the real motivation for the challenged strike(s). For discussion, see [@LaFave2020, sec. 22.3].

By law, evidence supporting a _Batson_ challenge may include not only the circumstances of the jury selection itself, but also data on peremptory challenges in past cases. Such information would be relevant, in any particular case, to whether (1) to conclude that the defendant has a prima facie case; (2) to believe a prosecutor from that office who has offered a race-neutral explanation for striking, say, an African-American juror in that case; and (3) to ultimately accept or reject a Batson challenge in any particular case.^[See, e.g., Miller-El v. Dretke, 545 U.S. 231, 253 (2005); Riley v. Taylor, 277 F.3d 261, 280-81, 283 (3d Cir. 2001); Cochran v. Herring, 43 F.3d 1404, 1412 (11th Cir. 1995); see also Flowers v. Mississippi, 139 S. Ct. 2228, 2243 (2019) (kinds of evidence to support Batson challenge includes "relevant history of the State's peremptory strikes in past cases")]

Some States have adopted variants of _Batson_ that reduces or eliminates the challenger's burden to establish a prima facie case^[E.g., State v. Holloway, 209 Conn. 636, 646 & n.4 (1989); State v. Chapman, 317 S.C. 302, 305-06 (1995); see also State v. Antwine, 743 S.W.2d 51, 64 (Mo. 1987) (trial judges must consider prosecutor's explanation to determine whether defendant established a prima facie case).]; as well as, in a few States, replaced a focus on the motive for strikes with a query into whether a hypothetical "objective observer" would find that race or ethnicity was a "factor" in the use of strikes.^[Wash. R. 37(e)].

@@

Here, we discuss how we developed a software prototype for attorneys to assess whether one can credibly infer bias, for Batson purposes, based on an attorney's pattern of peremptory challenges in a particular case, either alone or in combination with historical data on that attorney's pattern of strikes in past cases. In doing so, we highlight two distinctive contributions.

@@

Second, we implement the analysis as prototype software that, in principle, an attorney could use in real time during jury selection to assess how likely an attorney striking prospective jurors in violation of _Batson_.

@@

However, under this set up, the weight of non cognizable class is the reciprocal of the weight of cognizable class, which is not a desirable character. Using $b = log(w)$ could avoid this issue. For any given value of the bias parameter $b$, the probability of strike of a member from either class, or $Pr(\delta_{ij})$, is shown in equation \@ref(eq:model1)

@@

The posterior distribution does not have a closed form, thus Metropolis-Hasting algorithm is implemented to sample $b$ from the posterior distribution. Traceplots, Gelman-Rubin convergence diagnostic [@Gelman1992; @Gelman1998} and Geweke’s diagnostics [@Gewe1992] were used to measure the convergence. 
<!--

@@

We conduct simulation study to measure model performances under different scenarios. Different scenarios considers different bias and preference for cognizable class member, different amount of historical information and whether the preference changes over time. As for different preference for cognizable class, we generate data using seven different values of the bias parameter $b = (-3,-2,-1,0,1,2,3)$. As introduced in previous section, $b<0$ represents a preference for a juror within the cognizable class, and $b>0$ represents a bias against the cognizable class. Greater absolute value of $b$ represents greater preference or bias for the cognizable class. $b=0$ represents no bias. As for the amount of historical data, we consider three sizes of historical data, the same size as current data, double size of current data and triple size of current data. We also consider seven different values of the bias parameter for generating the historical data and allow the bias parameter of historical data to be different from current data. The different value of bias parameter from historical data to current data represents a change of preference for cognizable class.

When we use power prior incorporating the historical information, we can control the weight that we put on historical data by modifying the weight parameter $a$. In the simulation study, we also try different values of the weight parameter $a = (0.1,0.3,0.5,0.7,1)$ to show the influence of the weight parameter on modeling accuracy. We consider 735 scenarios in total, which is the combination of 7 different bias parameter of current data and of historical data, 3 different amount of historical data and 5 different values of the power prior weight parameter $a$. For each scenario, we generate 1000 datasets and fit the model on the 1000 generated dataset. We record the mean value of the posterior mean and credible intervals of the 1000 model fits for each scenario.

@@

Thus if the historical data is not reliable, we should set a small weight parameter on them. For example, [insert]

@@

The posterior probability distribution is estimated with R 4.0 [@R]. The app was built using the shiny package [@shiny].

>>>>

This paper demonstrates how to use contemporaneous and historical jury-selection data to estimate the probability that an attorney has used peremptory challenges to strike potential jurors because of juror race or gender, in violation of _Batson v. Kentucky_ (1986), and its progeny.

@@

In the US, once a group of individuals selected for jury-service appear in court as scheduled, this group is questioned by the parties' attorneys and the trial judge.  During this "voir dire" process, each prospective juror may be (1) excused by the trial judge for cause; (2) selected for the jury; or (3) dismissed upon the exercise by one of the parties' attorneys of a peremptory challenge. The key feature of a peremptory challenge is that, with it, a party declares a prospective juror ineligible ("strikes" that juror) for a seat on the jury, but without the burden of explaining why. Each side is assigned a number of peremptory challenges and may use them to strike any potential juror without explanation. The number of peremptory challenges varies by jurisdiction, as does peremptory-challenge procedure, including the order of strikes and the number of peremptory strikes a party may exercise at a time [@nscs-jury2018; @Williams2017].

@@

Under _Batson v. Kentucky_ (1986)_, however, a party's exercise of a peremptory challenge violates the Equal Protection Clause of the US Constitution if motivated by a prospective juror's race, gender, or certain other protected characteristics. When raising a _Batson_ challenge during jury selection, the challenging party must typically establish a "prima facie case", that is, enough evidence to infer that a party struck a juror because of such juror's race, gender, or other protected status. Once established, the challenged party bears the burden of producing one or more nondiscriminatory reasons for that strike.  Thereafter, the challenger bears the burden of persuading the trial judge that the race or gender, not the proffered nondiscriminatory reason, was more likely than not the real motivation for the challenged strike(s). For discussion, see [@LaFave2020, sec. 22.3]. Some States have adopted variants of _Batson_ that reduce or eliminates the challenger's burden to establish a prima facie case^[E.g., State v. Holloway, 209 Conn. 636, 646 & n.4 (1989); State v. Chapman, 317 S.C. 302, 305-06 (1995); see also State v. Antwine, 743 S.W.2d 51, 64 (Mo. 1987) (trial judges must consider prosecutor's explanation to determine whether defendant established a prima facie case).]; and, more recently, replaced a focus on the motive for strikes with a query into whether a hypothetical "objective observer" would find that race or ethnicity was a "factor" in the use of strikes.^[Wash. R. 37(e)].

By law, evidence supporting a _Batson_ challenge may include not only the circumstances of the jury selection itself, but also data on peremptory challenges in past cases. For example, in a _Batson_ challenge against a prosecutor in a criminal case, such historical data would be relevant to whether (1) to conclude that the defendant has a prima facie case; (2) to believe a prosecutor from that office who has offered a race-neutral explanation for striking, say, an African-American juror in that case; and (3) to ultimately accept or reject a Batson challenge in that case.^[See, e.g., Miller-El v. Dretke, 545 U.S. 231, 253 (2005); Riley v. Taylor, 277 F.3d 261, 280-81, 283 (3d Cir. 2001); Cochran v. Herring, 43 F.3d 1404, 1412 (11th Cir. 1995); see also Flowers v. Mississippi, 139 S. Ct. 2228, 2243 (2019) (kinds of evidence to support Batson challenge includes "relevant history of the State's peremptory strikes in past cases")]

@@

Here, we discuss how to assess whether one can credibly infer bias, for Batson purposes, based on an attorney's pattern of peremptory challenges in a particular case either alone or in combination with historical data on that attorney's pattern of strikes in past cases. In doing so, we make two distinctive contributions.

@@

Second, we implement the analysis as prototype software that an attorney could use in real time during jury selection to assess how likely an attorney striking prospective jurors in violation of _Batson_.

@@

However, under this set up, the weight of the non-cognizable class is the reciprocal of the weight of cognizable class, which is not a desirable character. Using $b = log(w)$ could avoid this issue. For any given value of the bias parameter $b$, the probability of strike of a member from either class, or $Pr(\delta_{ij})$, is shown in equation \@ref(eq:model1)

@@

The posterior distribution does not have a closed form.  Accordingly, we used the Metropolis-Hasting algorithm sample $b$ from the posterior distribution. Traceplots, Gelman-Rubin convergence diagnostic [@Gelman1992; @Gelman1998} and Geweke’s diagnostics [@Gewe1992] were used to measure the convergence. 
<!--

@@

We conducted a simulation study to measure model performance under different values of the bias parameter, different amount of historical information and whether the bias/preference changes over time. To do this, we generated data using seven different values of the bias parameter $b = {-3,-2,-1,0,1,2,3}$. As introduced above, $b<0$ represents a preference for a juror within the cognizable class, and $b>0$ represents a bias against the cognizable class. A greater absolute value of $b$ represents greater preference or bias for the cognizable class, while $b=0$ represents no bias. We also considered three sizes of historical data, the same size as current data, double size of current data and triple size of current data. We also considered seven different values of the bias parameter for generating the historical data and allow the bias parameter of historical data to be different from current data. The different value of bias parameter from historical data to current data represents a change of preference for cognizable class.

By using the power prior to incorporate the historical information, we can control the weight that we put on historical data by modifying the weight parameter $a$.  We may wish to exercise such control and $a < 1$ if some of the historical data is missing not at random. For example, for any given attorney, there may be only incomplete information available as to that attorney's exercise of peremptory challenges in previous trials. Possible reasons include limited access to the paper records containing such information, either due to rules governing who may access such records or the expense incurred in obtaining copies of such records. Standard jury-selection forms may not always require recording relevant information about prospective jurors (e.g., their race or gender), which may force researchers to combine such records with other data sources (e.g., voter registration rolls) by matching by name.  And even if standard forms or other recordkeeping rules require collecting all relevant information for each past trial, judges, attorneys, and court staff can vary in how much or how often they actually fill in all the required information.

Accordingly, in the simulation study, we tried different values of the weight parameter $a = (0.1,0.3,0.5,0.7,1)$ to show how that weight parameter affected modeling accuracy. We consider 735 scenarios in total, i.e., the combination of 7 different bias parameter of current data and of historical data, 3 different amount of historical data and 5 different values of the power prior weight parameter $a$. For each scenario, we generate 1000 datasets and fit the model on the 1000 generated dataset. We record the mean value of the posterior mean and credible intervals of one thousand model fits for each scenario.

@@

@@

The posterior probability distribution is estimated with `r R.version$version.string'[@R]. The app was built using the shiny package [@shiny].
